From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-30 Pi/paper/finite.tex
Copy-editing finite section

One way of thinking about presentations of $\Sn$ is via sorting algorithms, which use different primitive operations. A
sorting algorithm has to calculate a permutation of a list or a finite set, which satisfies the invariant of being a
sorted sequence, which means, the primitive operations of a sorting algorithm are able to generate all the permutations
on a given list. So, a chosen set of reversible operations in a sorting algorithm can be a good candidate for the
generators of a permutation group. For example, we could generate the symmetric group on $\Fin[n]$ by using generators
(primitive operations) that:

@@

We will show that it is possible to encode all $\PiLang$ combinators using adjacent transpositions
(in~\cref{sec:equivalence}). Group presentations are built by adding equations to a \emph{free group}.

@@

A group homomorphism between groups $G$ and $H$ is a function $f : G \to H$ between the underlying sets that preserves
the group structure. Giving a group homomorphism out of the free group is equivalent to giving a function out of the
generating set. This is the universal property of free groups, stemming from the free-forgetful adjunction between the
category of groups and sets.

@@

Following the universal-algebraic definition, in HoTT, we could use a naive higher inductive type to define the free
group, which enforces the group axioms by adding path constructors (see~\cref{def:free-group}). Using the induction
principle, we can easily verify the universal property. However, since this definition of $F(A)$ has lots of path
constructors corresponding to each group axiom, characterising its path space is difficult.

@@

A presentation of a group builds it by starting from the free group $F(A)$ and introducing additional equations that are
satisfied in the resulting group. For example, if we take $F(\unit) \defeq \mathbb{Z}$ and add an equation $1 + 1 = 0$,
the resulting group would be $\mathbb{Z}_2 \eqv \Aut[\Bool]$. Note that not all groups have finite (or computable)
presentations, and, a group can have any number of different presentations.

@@

The universal property of the above definition is similar to~\cref{prop:free-groups} except the relation has to be
preserved by the function mapping out of the generating set.

@@

Before, the only way to decide the equality of two elements in a group was to evaluate and check them on the nose, but
in a group presentation, this is reduced to deciding whether one word -- a representative of the equivalence class of
the group's elements, can be reduced to another word, using the group's relations. However, these equations are not
directed, so it is not always possible to construct a well-behaved rewriting system. In general, the word problem for
groups is undecidable.

\paragraph{Coxeter Presentation} To present the group $\Sn$, the primitive operations we use will be adjacent swaps.
When dealing with permutations on an $n + 1$-element set, there are $n$ adjacent transpositions -- transposition number
$k$ swaps elements at indices $k$ and $k+1$. Thus, the generating set is $\Fin[n]$. There are three relations that we're
going to specify for this presentation -- we visualise them as braid diagrams in~\cref{fig:coxeter-braid}.

@@

For this to work, we first need the system to have the termination property, meaning that there are no infinite
reductions. We observe that after throwing out reflexivity and symmetry, the right hand sides of the relations $\cox*$
are strictly smaller than the left hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$ (which is
well-founded). Thus, by directing the relation from left to right, we would get the termination property out of the box.
Second, we need the normal forms to be unique, so that we can get a normalisation function. This will be true if the
rewriting system is confluent -- meaning that all critical pairs, that is, terms with overlapping possible reduction
rules, have to converge. For example, in our system, the pair in~\cref{fig:critical-pairs-converging} converges.
Unfortunately, this is not true for all critical pairs -- an  example is in~\cref{fig:critical-pairs-non-converging},
where left and right endpoints are normal with respect to the $\cox*$ relation.

@@

Because of this counter-example, the relations have to be changed. In this section, we formally define a rewriting
system $(\List[\Fin[n]], \longcox)$, partially based on the Coxeter relations, and prove that it has the desired
properties of confluence and termination. We prove that the new relation defined by this system is equivalent, in a
technical sense, to the standard Coxeter relation $\cox*$. First, we need to define a function $n \downf k$.

@@

\noindent The result of this function is the sequence \([k + n - 1, k + n - 2, k + n - 3, \ldots, n]\), which is a
sequence of transpositions that moves the element at index $k + n$ left by $k$ places, shifting all the elements in
between one place right (see~\cref{fig:downf}). Then, the directed relation $\longcox$ is defined with the following
generators (inlining the congruence closure in $\longcox$, allowing arbitrary reductions inside the list).

@@

\noindent Constructors $\longcancel$ and $\longswap$ correspond directly to the appropriate constructors of $\cox$ and
can be visualised in the same way as before. The remaining constructor $\longbraid$ uses the $\downf$ function to
exchange the order of a long sequence of transpositions and a single transposition afterwards. For example, for $n = 0$
and $k = 3$, it allows for the reduction $[4, 3, 2, 1, 0, 4] \longcox [3, 4, 3, 2, 1, 0]$ (see~\cref{fig:longcox} --
note the distinction between wires, where numbers represent the values, and transpositions, where numbers represent
which wires are crossing).

>>>>

One way of thinking about presentations of $\Sn$ is via sorting algorithms, which use different primitive operations. A
sorting algorithm has to calculate a permutation of a list or a (listed) finite set, which satisfies the invariant of
being a sorted sequence, which means, the primitive operations of a sorting algorithm are able to generate all the
permutations on a given list. So, a chosen set of reversible operations in a sorting algorithm can be a good candidate
for the generators of a permutation group. For example, we could generate the symmetric group on $\Fin[n]$ by using
generators (primitive operations) that:

@@

We will show that it is possible to encode all $\PiLang$ combinators using adjacent transpositions
(in~\cref{sec:equivalence}).
%Group presentations are built by adding equations to a \emph{free group}.

@@

A group homomorphism between groups $G$ and $H$ is a function $f : G \to H$ between the underlying sets that preserves
the group structure. Giving a group homomorphism out of the free group is equivalent to giving a function out of the
generating set. This is the universal property of free groups, stemming from the free-forgetful adjunction between the
category of groups and sets. We say that $\eta_{A} : A \to F(A)$ is the free group on $A$ if it satisfies the following
universal property.

@@

Following the naive universal-algebraic definition, in HoTT, we could use a higher inductive type to define the free
group, which enforces the group axioms by adding path constructors (see~\cref{def:free-group}). Using the induction
principle, we can easily verify the universal property. However, since this definition of $F(A)$ has lots of path
constructors corresponding to each group axiom, characterising its path space is difficult.

@@

A presentation of a group builds it by starting from the free group $F(A)$ and introducing additional equations that are
satisfied in the resulting group. For example, if we take $F(\unit) \defeq \mathbb{Z}$ and add an equation $1 + 1 = 0$,
the resulting group would be $\mathbb{Z}_2 \eqv \Aut[\Bool]$. Note that not all groups have finite (or computable)
presentations, and, a group can have a number of different presentations.

@@

The universal property of the above definition of a presented group is similar to~\cref{prop:free-groups}, except the
function mapping out of the generating set also has to respect the relation.

@@

Before, the only way to decide the equality of two elements in a group was to evaluate and check them on the nose, but
in a group presentation, this is reduced to deciding whether one word -- a representative of the equivalence class of
the group's elements, can be reduced to another word, using the group's relations. However, these relations are not
directed, so it is not always possible to construct a well-behaved rewriting system. In general, the word problem for
groups is undecidable.

\paragraph{Coxeter Presentation} To present the group $\Sn$, the primitive operations we use will be adjacent swaps.
When dealing with permutations on an $n + 1$-element set, there are $n$ adjacent transpositions -- transposition number
$k$ swaps elements at indices $k$ and $k+1$. Thus, the generating set is $\Fin[n]$. There are three relations that we're
going to specify for this presentation -- we visualise them as braiding diagrams in~\cref{fig:coxeter-braid}.

@@

For this to work, we first need the system to have the termination property, meaning that there should be no infinite
reductions. We observe that after throwing out reflexivity and symmetry, the right hand sides of the relations $\cox*$
are strictly smaller than the left hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$ (which is
well-founded). Thus, by directing the relation from left to right, we would get the termination property out of the box.
Second, we need the normal forms to be unique, so that we can extract a normalisation function. This will be true if the
rewriting system is confluent -- all critical pairs, that is, terms with overlapping possible reduction rules, have to
converge. For example, in our system, the pair in~\cref{fig:critical-pairs-converging} converges. Unfortunately, this is
not true for all critical pairs -- an example is in~\cref{fig:critical-pairs-non-converging}, where left and right
endpoints are normal with respect to the $\cox*$ relation.

@@

Because of this counter-example, the relations have to be changed. In this section, we formally define a rewriting
system $(\List[\Fin[n]], \longcox)$, partially based on the Coxeter relations, and prove that it has the desired
properties of confluence and termination. We prove that the new relation defined by this system is equivalent, in a
technical sense (\cref{prop:coxlongcox}), to the standard Coxeter relation $\cox*$. First, we need to define a function
$\downf$.

@@

\noindent The result of computing $n \downf k$ is the sequence \([k + n - 1, k + n - 2, k + n - 3, \ldots, n]\), which
is a sequence of transpositions that moves the element at index $k + n$ left by $k$ places, shifting all the elements in
between one place right (see~\cref{fig:downf}). Then, the directed relation $\longcox$ is defined with the following
generators (inlining the congruence closure in $\longcox$ allows arbitrary reductions inside the list).

@@

\noindent Constructors $\longcancel$ and $\longswap$ correspond directly to the appropriate constructors of $\cox$ and
can be visualised in the same way as before. The remaining constructor $\longbraid$ uses the $\downf$ function to
exchange the order of a long sequence of transpositions and a single transposition afterwards. For example, for $n = 0$
and $k = 3$, it allows for the reduction $[4, 3, 2, 1, 0, 4] \longcox [3, 4, 3, 2, 1, 0]$ (see~\cref{fig:longcox} --
there is a distinction between wires, where numbers represent the values, and transpositions, where numbers represent
crossing wires).
